### 2014-08-18 doc/thesis/typedlua.tex
Fixing Chapter 3 to show types using concrete syntax instead of abstract syntax

\section{Types}
\label{sec:types}

Lua values can have one of eight tags:
\emph{nil}, \emph{boolean}, \emph{number}, \emph{string},
\emph{function}, \emph{table}, \emph{userdata}, and \emph{thread}.
In this section we present the types that Typed Lua introduces
to represent these tags.

\begin{figure}[!ht]
\textbf{Types}\\
\dstart
\begin{align*}
\textit{type} & ::= \; \textit{primarytype} \; [\texttt{`?'}]\\
\textit{primarytype} & ::= \; \textit{literaltype} \; | \;
  \textit{basetype} \; | \;
  \textbf{nil} \; | \;
  \textbf{value} \; | \;
  \textbf{any} \; | \;
  \textbf{self} \; | \;
  \textit{Name}\\
& | \; \textit{functiontype} \; | \;
  \textit{tabletype} \; | \;
  \textit{primarytype} \; \texttt{`|'} \; \textit{primarytype}\\
\textit{literaltype} & ::= \; \textbf{false} \; | \;
  \textbf{true} \; | \;
  \textit{Number} \; | \;
  \textit{String}\\
\textit{basetype} & ::= \; \textbf{boolean} \; | \;
  \textbf{number} \; | \;
  \textbf{string}\\
\textit{functiontype} & ::= \; \textit{tupletype} \; \texttt{`->'} \; \textit{rettypelist}\\
\textit{methodtype} & ::= \; \textit{tupletype} \; \texttt{`=>'} \; \textit{rettypelist}\\
\textit{tabletype} & ::= \; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}\\
\textit{tupletype} & ::= \; \texttt{`('} \; [typelist] \; \texttt{`)'}\\
\textit{typelist} & ::= \; \textit{type} \; \{\texttt{`,'} \; \textit{type}\} \; [\texttt{`*'}]\\
\textit{rettypelist} & ::= \; \textit{unionlist} \; [\texttt{`?'}]\\
\textit{unionlist} & ::= \; \textit{tupletype} \; | \;
  \textit{unionlist} \; \texttt{`|'} \; \textit{unionlist}\\
\textit{tabletypebody} & ::= \; \textit{type} \; | \; \textit{fieldtypelist}\\
\textit{fieldtypelist} & ::= \; [\textbf{const}] \; \textit{fieldtype} \; \{\texttt{`,'} \; [\textbf{const}] \; \textit{fieldtype}\}\\ 
\textit{fieldtype} & ::= \; \textit{keytype} \; \texttt{`:'} \; \textit{type}\\
\textit{keytype} & ::= \; \textit{literaltype} \; | \;
  \textit{basetype} \; | \;
  \textbf{value} \; | \;
  \textbf{any}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua types}
\label{fig:types}
\end{figure}

Figure \ref{fig:types} presents the concrete syntax of the
Typed Lua types.
We split Typed Lua types into two categories:
\emph{first-level types} and \emph{second-level types}.
We use first-level types to represent Lua values and
second-level types to type expression lists, multiple assignments,
and function applications.

Typed Lua uses subtyping to order types.
Any first-level type is a subtype of $\Value$.
The rest of the subtyping relationship is standard:
union types are supertypes of their respective literal types,
function types are related by contravariance on the input
and covariance on the output, table types have width subtyping,
with depth subtyping on $\Const$ fields, tuple and vararg types
are covariant.

Typed Lua uses consistent-subtyping to allow the dynamic type $\Any$
to interoperate between dynamically typed and statically typed code.
The dynamic type $\Any$ is a subtype of $\Value$, but neither a
supertype nor a subtype of any other type.
Our consistent-subtyping relationship follows the standards defined
by the gradual typing of objects \citep{siek2007objects,siek2013mutable}.
In practice, we can pass a value of the dynamic type anytime we want
a value of some other type, and can pass any value where where a
value of the dynamic type is expected, but these operations are tracked
by the type system, and the programmer can choose to be warned about them.

\section{Atomic types and Functions}

Typed Lua allows optional type annotations in variable and function
declarations.
Whenever it is possible to use local type inference, the compiler
assigns a more specific type to an unannotated declaration.
When it is not possible to use local type inference, the compiler
assigns the type $\Any$ to an unannotated declaration.
For instance, in function declarations, the compiler does not infer
a more specific type to unannotated input parameters, because,
in this case, global type inference often is undecidable in a
language with subtyping and overloading \citep{wells1999typability}.

In the following example we use type annotations in a function
declaration, and we do not use type annotations in a declaration of a
local variable:
\begin{verbatim}
    local function factorial (n:number):number
      if n == 0 then
        return 1
      else
        return n * factorial(n - 1)
      end
    end
    local x = 5
    print(factorial(x))
\end{verbatim}

Even though we did not add type annotations in the declaration of
the local variable \texttt{x}, the compiler uses local type inference to
assign the type $\Number$ to \texttt{x}, and this example compiles
without any warnings.
The compiler does not use the literal type $5$ instead of the base
type $\Number$ because this would prevent
programmers from assigning other numeric values to the variable
\texttt{x}.
The inference that we implement in Typed Lua is quite simple, as it
uses only the type of the local expression.

Typed Lua allows programmers to combine annotated code with
unannotated code, as we show in the following example:
\begin{verbatim}
    local function abs (n:number):number
      if n < 0 then
        return -n
      else
        return n
      end
    end

    local function dist (x, y)
      return abs(x - y)
    end
\end{verbatim}

The function \texttt{dist} receives two parameters of type $\Any$
and returns a value of type $\Number$.
The compiler assigns the dynamic type $\Any$ to the input
parameters of \texttt{dist} because they do not have type annotations,
and the compiler does not use global type inference, as we mentioned
previously.
Even though we did not annotated the return type of \texttt{dist},
the compiler is able to infer its return type because it is local
and not recursive.

We are using the consistent-subtyping relation of gradual
typing to check the interaction among the dynamic type and the other
types.
More precisely, we can call \texttt{abs} inside \texttt{dist} because
\texttt{x - y} is of type $\Any$, and the type $\Any$ is
consistent with the type $\Number$.
The type $\Any$ is not the union of all types, but it is a separate
type.

In this example, Typed Lua cannot guarantee that \texttt{dist} is never
going to call \texttt{abs} with a parameter that is not a number,
because in the semantics of Lua the minus operator can result in a
value that is not a number.
This is a typical example where run-time checks would ensure safety
between the interaction of dynamically typed and statically typed code.

Lua has first-class functions, and they have three peculiarities:
flexible arity, multiple return values, and multiple assignment.
In the first case, the number of arguments of a function call does
not need to match the arity of the function declaration, as Lua silently
drops extra arguments after evaluating them, or uses \texttt{nil} to
replace missing arguments.
In the second case, Lua function can return any number of values,
and this number of returned values may not be statically known.
In the third case, Lua has multiple assignment, and the semantics
of argument passing is the same of the multiple assignment
(or vice-versa), that is, calling a function is like doing a
multiple assignment where the left hand side is the parameter list
and the right hand side is the argument list.

Typed Lua uses \emph{second-level types} to encode functions with
flexible arity, multiple returns, and multiple assignment.
We call them second-level because these types do not correspond
to actual Lua values and we cannot use them to type variables or
parameters.
They are just an artifact of the interaction between the type system
and the semantics of Lua.

As we can see in Figure \ref{fig:typelang}, a second-level type is
a tuple of first-level types optionally ending in a variadic type,
or a union of these tuples.
A variadic type $T{*}$ is a generator for a sequence of values of
type $T \cup \Nil$.
Unions of tuples play an important role in functions that are
overloaded on the return type, together with \emph{projection types}.
We will explain union types and projection types in more detail
in the next section.

To match the semantics of Lua function calls, the default operation
mode of Typed Lua adds a variadic tail to the parts of a function
type whenever the programmer does not specify one.
In the previous examples, the type of both \texttt{factorial} and
\texttt{abs} is
$\Number \times \Value{*} \rightarrow \Number \times \Nil{*}$,
and the type of \texttt{dist} is
$\Any \times \Any \times \Value{*} \rightarrow \Number \times \Nil{*}$.
If we call \texttt{abs} with extra arguments, Typed Lua silently ignores
them, as the type signature lets \texttt{abs} receive any number of
extra arguments.
If we call \texttt{abs} with fewer arguments, Typed Lua uses $\Nil$ to
replace the missing arguments and in the case of \texttt{abs} generates
a warning, as $\Nil$ is neither a subtype of $\Number$ nor consistent
with $\Number$.

Typed Lua also provides a strict operation mode, and programmers can
use this mode if they want to catch arity mismatch of function calls.
In the strict mode, Typed Lua does not add a variadic tail to the
parts of a function type unless the programmer explicitly declares one.
In the previous examples, the type of both \texttt{factorial} and
\texttt{abs} is $\Number \rightarrow \Number$, and the type of
\texttt{dist} is $\Any \times \Any \rightarrow \Number$.
If we call \texttt{abs} with either extra or missing arguments,
Typed Lua generates warnings about arity mismatch.

@@

In the default mode, the type of \texttt{multiple} is
$\Value{*} \rightarrow \Number \times \String \times \Nil{*}$,
and the type of \texttt{sum} is
$\Number \times \Number \times \Value{*} \rightarrow \Number \times \Nil{*}$.
In the right hand side of the multiple assignment, only the first
value produced by the first call to multiple gets used,
so the type of the right hand side is
$\Number \times \Number \times \String \times \Nil{*}$,
and the type assigned to \texttt{x}, \texttt{y}, and \texttt{z}
are respectively $\Number$, $\Number$, and $\String$.
The same tuple is assigned to the call to \texttt{sum} and it
compiles without errors, as the first two components of the tuple
are consistent with the types of the parameters, and the other
components are consistent with \texttt{value}.

In the strict mode, the type of \texttt{multiple} is
$\Void \rightarrow \Number \times \String$,
and the type of \texttt{sum} is $\Number \times \Number \rightarrow \Number$.
Like in the default mode, only the first value produced by the
first call to multiple gets used in the right hand side of the
multiple assignment;
and the type of the right hand side is also
$\Number \times \Number \times \String \times \Nil{*}$,
because the strict mode affects only the type of function declarations
and multiple assignments that appear in function calls.
Unlike in the default mode, the type of the call to \texttt{sum} is
$\Number \times \Number \times \String$, and does not include the
variadic tail to catch arity mismatch.
Thus, this example compiles with error because the call to \texttt{sum}
passes an unexpected extra argument of type $\String$.

@@

Optional values are unions of some type $t$ and $\Nil$, and Typed Lua
includes the syntactic sugar \texttt{t?} to represent them because
they appear quite often.
The concrete syntax \texttt{t?} is syntactic sugar for \texttt{t|nil}
that maps to the type $T\;\cup\;\Nil$.
Optional values can appear when a function has optional parameters
and when the program reads a value from an array or a map.
The following example shows a function that has an optional parameter:
\begin{verbatim}
    local function message (name:string, greeting:string?)
      local greeting = greeting or "Hello "
      return greeting .. name
    end

    print(message("Typed Lua"))         -- prints Typed Lua
    print(message("Typed Lua", "Hi "))  -- prints Hi Typed Lua
\end{verbatim}

Although the parameter \texttt{greeting} is optional, and has type
$\String\;\cup\;\Nil$, the concatenation does not generate a type error
because we used the short-circuiting \texttt{or} operator to declare a
new variable \texttt{greeting} that is guaranteed to have type $\String$. 
In Lua, only the values \texttt{nil} and \texttt{false} represent a
false condition, so programmers often use the \texttt{or} operator as a
common idiom for giving a default value to an optional parameter. 
Typed Lua uses a typing rule to encode this idiom: 
if the left hand side of \texttt{or} has type $T\;\cup\;\Nil$ and
the right hand side has type $T$ then the \texttt{or} expression has
type $T$.

In fact, we do need to declare a new variable \texttt{greeting} that
shadows the optional parameter:
\begin{verbatim}
    local function message (name:string, greeting:string?)
      greeting = greeting or "Hello "
      return greeting .. name
    end
\end{verbatim}

Typed Lua allows the assignment $x\;=\;x\;\mathbf{or}\;e$ to refine
the type of $x$ from $T\;\cup\;\Nil$ to $T$ if it matches three restrictions: 
the type of $e$ is a subtype of $T$, the variable $x$ is local to the
current function, and it is not being assigned in another function.
The refinement only affects the type of $x$ in the remainder of the
current scope.
In the case of \texttt{greeting}, the assignment refines its type
from $\String\;\cup\;\Nil$ to $\String$.

@@

Typed Lua has a small set of \texttt{type} predicates that allows
programmers to constrain the type of a local variable inside a condition.
This example uses the \texttt{type(x) == "string"} predicate that
constrains the type of $x$ from $T\;\cup\;\String$ to $\String$ when
the predicate is true and $T$ otherwise.
This is a simplified form of \emph{flow typing} \citep{guha2011tlc,tobin-hochstadt2010ltu}.
As with \texttt{or}, the variable must be local to the function,
and it is not being assigned in another function.

The \texttt{type} predicates can only discriminate based on tags,
so they are limited on the kinds of unions that they can discriminate.
For instance, the predicates can discriminate a union that combines
a table type with a base type, or a table type with a function type,
or two base types, but they cannot discriminate a union that combines
two different table types, or two different table types.

@@

Typed Lua also includes a syntactic sugar for this idiom:
we can annotate the return type of \texttt{idiv} with
\texttt{(number, number)?} to denote the same union.
The parentheses are always necessary in this case because
\texttt{t?} is syntactic sugar for \texttt{t|nil},
while \texttt{(t)?} is syntactic sugar for \texttt{(t)|(nil, string)}.

The type of idiv, in the default mode, is
$\Number \times \Number \times \Value{*} \rightarrow
(\Number \times \Number \times \Nil{*}) \cup
(\Nil \times \String \times \Nil{*})$
A typical client of this function would use it as follows:
\begin{verbatim}
    local q, r = idiv(a, b)
    if q then
      print(a == b * q + r)
    else
      print("ERROR: " .. r)
    end
\end{verbatim}

When Typed Lua finds a union of tuples in the right hand side of
a declaration, it stores this union in a special type environment
with a fresh name and assigns \emph{projection types} to the
variables that appear unannotated in the left hand side of the declaration.
In our example, Typed Lua assigns the tuple
$(\Number \times \Number \times \Nil{*}) \cup (\Nil \times \String \times \Nil{*})$
to a type variable $X$.
Thus, variable \texttt{q} gets the type $X_{1}$
that is equivalent to $\Number\;\cup\;\Nil$, 
and variable \texttt{r} gets the type $X_{2}$
that is equivalent to $\Number\;\cup\;\String$.

Projection types do not appear in the subtyping and consistent-subtyping
relations because Typed Lua actually replaces them with the union of the
corresponding component of each tuple.
If a variable with a projection type appears in a \texttt{type} predicate,
it discriminates against all tuples in the union.
In our example, $X$ is
$(\Number \times \Number \times \Nil{*}) \cup (\Nil \times \String \times \Nil{*})$
outside of the \texttt{if} statement,
but $\Number \times \Number \times \Nil{*}$ inside the \texttt{then} block,
and $\Nil \times \String \times \Nil{*}$ inside the \texttt{else} block.
Thus, variable \texttt{q} has type $\Number$ and variable \texttt{r} also
has type $\Number$ inside the \texttt{then} block;
and variable \texttt{q} has type $\Nil$ while variable \texttt{r}
has type $\String$ inside the \texttt{else} block.

We could also have used \texttt{type(q) == "number"} or
even \texttt{type(r) == "number"} as the predicate of our example,
as both predicates would produce the same result.
However, the form that appears in our example is much more succint and idiomatic.
We can also use projection types to write overloaded functions where
the type of an input parameter depends on the type of another input parameter.

@@

This example shows that we cannot write a function that is
guaranteed to return a number when we pass two numbers and
guaranteed to return a string when we pass two strings.
Even though we can use intersection types to express the type
of this function as
$(\Number \times \Number \rightarrow \Number) \cap (\String \times \String \rightarrow \String)$,
and thefore solving the limitation,
intersection types require more sophisticated flow typing to
check whether a function has this type,
and we are still working on this problem.

@@

Tables are the only mechanism that Lua has to build data structures;
they are associative arrays where any value (except \texttt{nil})
can be used as a key.
Programmers can use tables to represent tuples, arrays (dense or sparse),
records, graphs, modules, objects, etc.
Lua has syntactic sugar for indexing tables as records:
\texttt{t.k} is syntactic sugar for \texttt{t["k"]}, and means that
we are accessing field \texttt{k} from table \texttt{t}. 
In this section, we show how Typed Lua encodes tables as maps,
arrays, and records.

Table types are the only framework that Typed Lua has to represent
the different uses that a Lua table has.
A table type $\{K_{1}:T_{1}, ..., K_{n}:T_{n}\}$ represents a map
from values of type $K_{i}$ to values of type $T_{i}$.

Although Typed Lua has just one framework to represent tables,
it has three different syntaxes for defining the type of a table.
The first syntax defines table types for maps:
it is written \texttt{\{k:t\}},
and maps to the table type $\{K:T\;\cup\;\Nil\}$.
This table type represents a map that maps values of type
$K$ to values of type $T\;\cup\;\Nil$,
as Lua returns $\Nil$ when we use a non-existing key to index a table.
The second syntax defines table types for arrays:
it is written \texttt{\{t\}},
and maps to the table type $\{\Number:T\;\cup\;\Nil\}$.
This table type is just a syntactic sugar to the definition of a
map that maps values of type $\Number$ to values of type $T\;\cup\;\Nil$.
The third syntax defines table types for records:
it is written \texttt{\{k1:t1, ..., kn:tn\}},
and maps to the table type $\{K_{1}:T_{1}, ..., K_{n}:T_{n}\}$.
This table type represents a record that maps the literal types
$K_{i}$\texttt{, ..., }$K_{n}$ to values of types
$T_{i}$\texttt{, ..., }$T_{n}$, where $i$ may vary from one to $n$.

>>>>

\section{Optional type annotations}

Lua values can have one of eight tags:
\emph{nil}, \emph{boolean}, \emph{number}, \emph{string},
\emph{function}, \emph{table}, \emph{userdata}, and \emph{thread}.
Typed Lua includes types for the first six.
Typed Lua also includes a syntactical extension that programmers can use
to define the types of \emph{userdata}, and uses it to define the type of \emph{thread}. 
In this section, we present the Typed Lua types that may appear on annotations.
We explain all Typed Lua types and syntactical extensions along this chapter.

\begin{figure}[!ht]
\textbf{Type annotations}\\
\dstart
\begin{align*}
\textit{type} & ::= \; \textit{primarytype} \; [\texttt{`?'}]\\
\textit{primarytype} & ::= \; \textit{literaltype} \; | \;
  \textit{basetype} \; | \;
  \textbf{nil} \; | \;
  \textbf{value} \; | \;
  \textbf{any} \; | \;
  \textbf{self} \; | \;
  \textit{Name}\\
& | \; \textit{functiontype} \; | \;
  \textit{tabletype} \; | \;
  \textit{primarytype} \; \texttt{`|'} \; \textit{primarytype}\\
\textit{literaltype} & ::= \; \textbf{false} \; | \;
  \textbf{true} \; | \;
  \textit{Number} \; | \;
  \textit{String}\\
\textit{basetype} & ::= \; \textbf{boolean} \; | \;
  \textbf{number} \; | \;
  \textbf{string}\\
\textit{functiontype} & ::= \; \textit{tupletype} \; \texttt{`->'} \; \textit{rettupletype}\\
\textit{methodtype} & ::= \; \textit{tupletype} \; \texttt{`=>'} \; \textit{rettupletype}\\
\textit{tabletype} & ::= \; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}\\
\textit{tupletype} & ::= \; \texttt{`('} \; [typelist] \; \texttt{`)'}\\
\textit{typelist} & ::= \; \textit{type} \; \{\texttt{`,'} \; \textit{type}\} \; [\texttt{`*'}]\\
\textit{rettupletype} & ::= \; \textit{unionlist} \; [\texttt{`?'}]\\
\textit{unionlist} & ::= \; \textit{tupletype} \; | \;
  \textit{unionlist} \; \texttt{`|'} \; \textit{unionlist}\\
\textit{tabletypebody} & ::= \; \textit{type} \; | \; \textit{fieldtypelist}\\
\textit{fieldtypelist} & ::= \; [\textbf{const}] \; \textit{fieldtype} \; \{\texttt{`,'} \; [\textbf{const}] \; \textit{fieldtype}\}\\ 
\textit{fieldtype} & ::= \; \textit{keytype} \; \texttt{`:'} \; \textit{type}\\
\textit{keytype} & ::= \; \textit{literaltype} \; | \;
  \textit{basetype} \; | \;
  \textbf{value} \; | \;
  \textbf{any}\\
\textit{rettype} & ::= \; \textit{type} \; | \; \textit{rettupletype}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua types}
\label{fig:types}
\end{figure}

Figure \ref{fig:types} presents the concrete syntax of the
Typed Lua types in extended BNF.
We classify Typed Lua types into two categories:
\emph{first-level types} and \emph{second-level types}.
First-level types consist of \emph{type} and represent Lua values,
while second-level types consist of either \emph{tupletype} or
\emph{rettype} and represent the type of expression lists,
multiple assignments, and function applications.
First-level types include literal types, base types, the type \texttt{nil},
the top type \texttt{value}, the dynamic type \texttt{any},
the self type \texttt{self}, named types, function types, table types,
and union types.
Second-level types include the type void \texttt{()}, first-level types,
vararg types, tuple types, and union of tuple types.

Typed Lua uses subtyping to order types.
Any first-level type is a subtype of \texttt{value}.
The rest of the subtyping relationship is standard:
union types are supertypes of their parts,
\texttt{boolean}, \texttt{number}, and \texttt{string} are supertypes
of their respective literal types,
function types are related by contravariance on the input
and covariance on the output, table types have width subtyping,
with depth subtyping on \texttt{const} fields, tuple and vararg types
are covariant.

Typed Lua uses consistent-subtyping to check the interaction among the
dynamic type \texttt{any} and other types.
The dynamic type \texttt{any} is a subtype of \texttt{value}, but neither
a supertype nor a subtype of any other type.
Our consistent-subtyping relationship follows the standards defined
by the gradual typing of objects \citep{siek2007objects,siek2013mutable}.
In practice, we can pass a value of the dynamic type anytime we want
a value of some other type, and can pass any value where a
value of the dynamic type is expected, but the compiler tracks these
operations, and the programmer can choose to be warned about them.

Typed Lua allows optional type annotations in variable and function
declarations.
Whenever it is possible to use local type inference, the compiler
assigns a more specific type to an unannotated declaration.
When it is not possible to use local type inference, the compiler
assigns the type \texttt{any} to an unannotated declaration.
For instance, in function declarations, the compiler does not infer
a more specific type to unannotated input parameters, because,
in this case, global type inference often is undecidable in a
language with subtyping and overloading \citep{wells1999typability}.

In the following example, we use type annotations in a function
declaration, and we do not use type annotations in a declaration of a
local variable:
\begin{verbatim}
    local function factorial (n:number):number
      if n == 0 then
        return 1
      else
        return n * factorial(n - 1)
      end
    end
    local x = 5
    print(factorial(x))
\end{verbatim}

Even though we did not add type annotations in the declaration of
the local variable \texttt{x}, the compiler uses local type inference to
assign the type \texttt{number} to \texttt{x}, and this example compiles
without any errors.
The compiler does not use the literal type \texttt{5},
instead of the base type \texttt{number}, because this would prevent
programmers from assigning other numeric values to the variable
\texttt{x}.
The inference that we implement in Typed Lua is quite simple, as it
uses only the type of the local expression.

Typed Lua allows programmers to combine statically typed code with
dynamically typed code, as we show in the following example:
\begin{verbatim}
    local function absolute (n:number):number
      if n < 0 then
        return -n
      else
        return n
      end
    end

    local function distance (x, y)
      return abs(x - y)
    end
\end{verbatim}

The function \texttt{distance} receives two parameters of type \texttt{any}
and returns a value of type \texttt{number}.
The compiler assigns the dynamic type \texttt{any} to the input
parameters of \texttt{distance} because they do not have type annotations,
and the compiler does not use global type inference, as we mentioned
previously.
Even though we did not annotated the return type of \texttt{distance},
the compiler is able to infer its return type because it is local
and not recursive.

In this example, Typed Lua cannot guarantee that \texttt{distance} is never
going to call \texttt{absolute} with a parameter that is not a number,
because in the semantics of Lua the minus operator can result in a
value that is not a number.
More precisely, we can call \texttt{absolute} inside \texttt{distance}
because the expression \texttt{x - y} has type \texttt{any},
and it is consistent with type \texttt{number}.
Still, the dynamic type \texttt{any} may be hiding a value of a type
that is not a number, making the dynamically typed code crash the
statically typed code.
This is a typical example where run-time checks would ensure safety
between the interaction of dynamically typed and statically typed code.

\section{Functions}

Lua has first-class functions, but they have some peculiarities.
First, the number of arguments of a function call does
not need to match the arity of the function declaration, as Lua silently
drops extra arguments after evaluating them, or uses \texttt{nil} to
replace missing arguments.
Second, functions can return any number of values,
and this number of returned values may not be statically known.
Third, Lua has multiple assignment, and the semantics
of argument passing is the same of the multiple assignment
(or vice-versa), that is, calling a function is like doing a
multiple assignment where the left hand side is the parameter list
and the right hand side is the argument list.

Typed Lua uses second-level types to encode function types,
and to preserve these peculiarities.
We call them second-level because these types do not correspond
to actual Lua values and we cannot use them to type variables or
parameters.
They are just an artifact of the interaction between the type system
and the semantics of Lua.

As we can see in Figure \ref{fig:types}, second-level types consist
of either \emph{tupletype} or \emph{rettupletype}.
A second-level type is the type void \texttt{()},
a tuple of first-level types optionally ending in a variadic type,
or a union of these tuples.
A variadic type \texttt{t*} is a generator for a sequence of values
of the union type \texttt{t|nil}.
Union of tuples may appear only on the return type of functions,
and together with \emph{projection types} play an important role
in functions that are overloaded on the return type.
We will explain union types and projection types in more detail
in the next section.

To match the semantics of Lua function calls, the default operation
mode of Typed Lua adds a variadic tail to the parts of a function
type whenever the programmer does not specify one.
In the previous examples, the type of both \texttt{factorial} and
\texttt{absolute} is \texttt{(number, value*) -> (number, nil*)},
and the type of \texttt{distance} is \texttt{(any, any, value*) -> (number, nil*)}.
If we call \texttt{absolute} with extra arguments, Typed Lua silently ignores
them, as the type signature lets \texttt{absolute} receive any number of
extra arguments.
If we call \texttt{absolute} with fewer arguments, Typed Lua uses \texttt{nil} to
replace the missing arguments and in the case of \texttt{absolute} generates
an error, as \texttt{nil} is neither a subtype of \texttt{number} nor consistent
with \texttt{number}.

Typed Lua also provides a strict operation mode, and programmers can
use this mode if they want to catch arity mismatch of function calls.
In the strict mode, Typed Lua does not add a variadic tail to the
parts of a function type unless the programmer explicitly declares one.
In the previous examples, the type of both \texttt{factorial} and
\texttt{absolute} is \texttt{(number) -> (number)},
and the type of \texttt{distance} is \texttt{(any, any) -> (number)}.
If we call \texttt{absolute} with either extra or missing arguments,
Typed Lua generates errors about arity mismatch.

@@

In the default mode, the type of \texttt{multiple} is
\texttt{(value*) -> (number, string, nil*)},
and the type of \texttt{sum} is \texttt{(number, number, value*) -> (number, nil*)}.
In the right hand side of the multiple assignment, only the first
value produced by the first call to multiple gets used,
so the type of the right hand side is
\texttt{(number, number, string, nil*)},
and the type assigned to \texttt{x}, \texttt{y}, and \texttt{z}
are respectively \texttt{number}, \texttt{number}, and \texttt{string}.
The same tuple is assigned to the call to \texttt{sum} and it
compiles without errors, as the first two components of the tuple
are consistent with the types of the parameters, and the other
components are consistent with \texttt{value}.

In the strict mode, the type of \texttt{multiple} is \texttt{() -> (number, string)},
and the type of \texttt{sum} is \texttt{(number, number) -> (number)}.
Like in the default mode, only the first value produced by the
first call to multiple gets used in the right hand side of the
multiple assignment;
and the type of the right hand side is also
\texttt{(number, number, string, nil*)},
because the strict mode affects only the type of function declarations
and multiple assignments that appear in function calls.
Unlike in the default mode, the type of the call to \texttt{sum} is
\texttt{(number, number, string)}, and does not include the
variadic tail to catch arity mismatch.
Thus, this example compiles with error because the call to \texttt{sum}
passes an unexpected extra argument of type \texttt{string}.

@@

Optional values are unions of some type \texttt{t} and \texttt{nil},
and Typed Lua includes the syntactic sugar \texttt{t?} to represent
them because they appear quite often.
The concrete syntax \texttt{t?} is syntactic sugar for \texttt{t|nil}.
Optional values can appear when a function has optional parameters
and when the program reads a value from an array or a map.
The following example shows a function that has an optional parameter:
\begin{verbatim}
    local function message (name:string, greeting:string?)
      local greeting = greeting or "Hello "
      return greeting .. name
    end

    print(message("Typed Lua"))         -- Typed Lua
    print(message("Typed Lua", "Hi "))  -- Hi Typed Lua
\end{verbatim}

Although the parameter \texttt{greeting} is optional, and has type
\texttt{string|nil}, the concatenation does not generate a type error
because we used the short-circuiting \texttt{or} operator to declare a
new variable \texttt{greeting} that is guaranteed to have type \texttt{string}. 
In Lua, only the values \texttt{nil} and \texttt{false} represent a
false condition, so programmers often use the \texttt{or} operator as a
common idiom for giving a default value to an optional parameter. 
Typed Lua uses a typing rule to encode this idiom: 
if the left hand side of \texttt{or} has type \texttt{t|nil} and
the right hand side has type \texttt{t} then the \texttt{or} expression has
type \texttt{t}.

In fact, we do not need to declare a new variable \texttt{greeting} that
shadows the optional parameter:
\begin{verbatim}
    local function message (name:string, greeting:string?)
      greeting = greeting or "Hello "
      return greeting .. name
    end
\end{verbatim}

Typed Lua allows the assignment \texttt{x = x or e} to refine
the type of \texttt{x} from \texttt{t|nil} to \texttt{t}
if it matches the following restrictions: 
the type of \texttt{e} is a subtype of \texttt{t},
and the variable \texttt{x} is local to the current function
and it is not being assigned in another function.
The refinement only affects the type of \texttt{x} in the remainder
of the current scope.
In the case of \texttt{greeting}, the assignment refines its type
from \texttt{string|nil} to \texttt{string}.

@@

Typed Lua has a small set of \texttt{type} predicates that allows
programmers to constrain the type of a local variable inside a condition.
This example uses the \texttt{type(x) == "string"} predicate that
constrains the type of \texttt{x} from \texttt{t|string} to
\texttt{string} when the predicate is true and \texttt{t} otherwise.
This is a simplified form of \emph{flow typing} \citep{guha2011tlc,tobin-hochstadt2010ltu}.
As with \texttt{or}, the variable must be local to the function
and it is not being assigned in another function.

The \texttt{type} predicates can only discriminate based on tags,
so they are limited on the kinds of unions that they can discriminate.
For instance, the predicates can discriminate a union that combines
a table type with a base type, or a table type with a function type,
or two base types, but they cannot discriminate a union that combines
two different table types, or two different function types.

@@

Typed Lua also includes a syntactic sugar for this idiom:
we can annotate the return type of \texttt{idiv} with
\texttt{(number, number)?} to denote the same union.
The parentheses are always necessary in this case, because
\texttt{t?} is syntactic sugar for \texttt{t|nil},
while \texttt{(t)?} is syntactic sugar for \texttt{(t)|(nil, string)}.

A typical client of this function would use it as follows:
\begin{verbatim}
    local q, r = idiv(a, b)
    if q then
      print(a == b * q + r)
    else
      print("ERROR: " .. r)
    end
\end{verbatim}

When Typed Lua finds a union of tuples in the right hand side of
a declaration, it stores this union in a special type environment
with a fresh name and assigns \emph{projection types} to the
variables that appear unannotated in the left hand side of the declaration.
In our example, Typed Lua assigns the union of tuples
\texttt{(number, number, nil*)|(nil, string, nil*)} to a type variable $X$.
Thus, variable \texttt{q} gets the type $X_{1}$
that is equivalent to \texttt{number|nil},
and variable \texttt{r} gets the type $X_{2}$
that is equivalent to \texttt{number|string}.

Projection types do not appear in the subtyping and consistent-subtyping
relations because Typed Lua actually replaces them with the union of the
corresponding component of each tuple.
If a variable with a projection type appears in a \texttt{type} predicate,
it discriminates against all tuples in the union.
In our example, $X$ is \texttt{(number, number, nil*)|(nil, string, nil*)}
outside of the \texttt{if} statement,
but \texttt{(number, number, nil*)} inside the \texttt{then} block,
and \texttt{(nil, string, nil*)} inside the \texttt{else} block.
Thus, variable \texttt{q} has type \texttt{number} and variable \texttt{r} also
has type \texttt{number} inside the \texttt{then} block;
and variable \texttt{q} has type \texttt{nil} while variable \texttt{r}
has type \texttt{string} inside the \texttt{else} block.

We could also have used \texttt{type(q) == "number"} or
even \texttt{type(r) == "number"} as the predicate of our example,
as both predicates would produce the same result.
However, the form that appears in our example is much more succint and idiomatic.

@@

This example shows that we cannot write a function that is
guaranteed to return a number when we pass two numbers and
guaranteed to return a string when we pass two strings.
Even though we can use intersection types to express the type
of this function as
\texttt{(number, number, value*) -> (number, nil*) \string&
(string, string, value*) -> (string, nil*)},
and thefore solving this limitation,
intersection types require more sophisticated flow typing to
check whether a function has this type,
and we are still working on this problem.

@@

Tables are the only mechanism that Lua has to build data structures;
they are associative arrays where any value (except \texttt{nil})
can be used as a key.
Programmers can use tables to represent tuples, arrays (dense or sparse),
records, graphs, modules, objects, etc.
Lua has syntactic sugar for indexing tables as records:
\texttt{t.k} is syntactic sugar for \texttt{t["k"]}, and means that
we are accessing the field \texttt{k} from table \texttt{t}. 
In this section, we show how Typed Lua encodes tables as maps,
arrays, and records.

Table types are the only framework that Typed Lua has to represent
the different uses that a Lua table has.
A table type $\texttt{\{k}{_\texttt{1}}\texttt{:t}_{\texttt{1}}
\texttt{, ..., k}{_\texttt{n}}\texttt{:t}_{\texttt{n}}\texttt{\}}$ represents a map
from values of type $\texttt{k}_{\texttt{i}}$ to values of type
$\texttt{t}_{\texttt{i}}$.
